---
output: html_document
---
# Practical Machine Learning Final Project
Author:  Felix Ling, Date: September 3, 2016

The main issue in creating a model was that there were a very large number of variables, and thus building the more complex models like Random Forest took prohibitively long and sometimes produced warnings (e.g., "variable 'kurtosis_roll_belt' is not a factor").

Upon inspection, there were a number of variables that only assumed one of two possible values: blank and "div/0" (an undefined value resulting in a division by 0). These were:

kurtosis_yaw_belt, skewness_yaw_belt, kurtosis_yaw_dumbbell, skewness_yaw_dumbbell, amplitude_yaw_dumbbell, kurtosis_yaw_forearm, skewness_yaw_forearm, amplitude_yaw_forearm

Likewise, these variables were blank, "div/0," or a handful of other values:

amplitude_yaw_belt, kurtosis_roll_arm, skewness_roll_arm, kurtosis_picth_arm, skewness_pitch_arm, kurtosis_yaw_arm, skewness_yaw_arm, kurtosis_roll_dumbbell, skewness_roll_dumbbell, kurtosis_picth_dumbbell, skewness_pitch_dumbbell, kurtosis_roll_belt, kurtosis_roll_forearm, skewness_roll_forearm, kurtosis_picth_forearm, skewness_pitch_forearm

It isn't that surprising that these variables were not that useful because they describe the distribution of other variables.

```{r include=FALSE}
library ("labeling", lib.loc="../../Documents/R/packages/")
library ("ggplot2", lib.loc="../../Documents/R/packages/")
library ("e1071", lib.loc="../../Documents/R/packages/")
library ("caret", lib.loc="../../Documents/R/packages/")
load ("data/PmlSub48Rf.RData")

# Subset the training set.
PmlSub <- Pml[8:dim(Pml)[2]]

# Filter out skewness and kurtosis and other useless variables.
PmlSub <- PmlSub[,-grep("kurt", colnames (PmlSub))]
PmlSub <- PmlSub[,-grep("skew", colnames (PmlSub))]
```

Next, viewing the structure of the dataframe shows that quite a few variables are mostly made up of NAs, meaning that the observation was not available. To keep this report concise, this command only shows the first 12.

```{r}
summary (PmlSub[1:12])
```

Refer to the summaries of max_roll_belt, max_pictch_belt, min_roll_belt, for examples with mostly NA's.

Overall, these variables had 19216 NA's out of a total of 19622 observations, meaning that most of the data was not available:

max_roll_belt, max_picth_belt, min_roll_belt, min_pitch_belt, amplitude_pitch_belt, var_total_accel_belt, avg_roll_belt, stddev_roll_belt, var_roll_belt, avg_pitch_belt, stddev_pitch_belt, var_pitch_belt, avg_yaw_belt, var_accel_arm, avg_roll_arm, stddev_roll_arm, var_roll_arm, avg_pitch_arm, stddev_pitch_arm, var_pitch_arm, avg_yaw_arm, stddev_yaw_arm, var_yaw_arm, max_roll_arm, max_picth_arm, max_yaw_arm, min_roll_arm, min_pitch_arm, min_yaw_arm, amplitude_roll_arm, amplitude_pitch_arm, amplitude_yaw_arm, max_roll_dumbbell, max_picth_dumbbell, min_roll_dumbbell, min_pitch_dumbbell, amplitude_roll_dumbbell, amplitude_pitch_dumbbell, var_accel_dumbbell, avg_roll_dumbbell, stddev_roll_dumbbell, var_roll_dumbbell, avg_pitch_dumbbell, stddev_pitch_dumbbell, var_pitch_dumbbell, avg_yaw_dumbbell, stddev_yaw_dumbbell, var_yaw_dumbbell, max_roll_forearm, max_picth_forearm, min_roll_forearm, min_pitch_forearm, amplitude_roll_forearm, amplitude_pitch_forearm, var_accel_forearm, avg_roll_forearm, stddev_roll_forearm, var_roll_forearm, avg_pitch_forearm, stddev_pitch_forearm, var_pitch_forearm, avg_yaw_forearm, stddev_yaw_forearm, var_yaw_forearm

```{r include=FALSE}
library ("reshape", lib.loc="../../Documents/R/packages/")

# Remove the ones with 19612 NAs.
PmlSub <- subset( PmlSub, select = -c(max_roll_belt, max_picth_belt, min_roll_belt, min_pitch_belt, amplitude_pitch_belt, var_total_accel_belt, avg_roll_belt, stddev_roll_belt, var_roll_belt, avg_pitch_belt, stddev_pitch_belt, var_pitch_belt, avg_yaw_belt, var_accel_arm, avg_roll_arm, stddev_roll_arm, var_roll_arm, avg_pitch_arm, stddev_pitch_arm, var_pitch_arm, avg_yaw_arm, stddev_yaw_arm, var_yaw_arm, max_roll_arm, max_picth_arm, max_yaw_arm, min_roll_arm, min_pitch_arm, min_yaw_arm, amplitude_roll_arm, amplitude_pitch_arm, amplitude_yaw_arm, max_roll_dumbbell, max_picth_dumbbell, min_roll_dumbbell, min_pitch_dumbbell, amplitude_roll_dumbbell, amplitude_pitch_dumbbell, var_accel_dumbbell, avg_roll_dumbbell, stddev_roll_dumbbell, var_roll_dumbbell, avg_pitch_dumbbell, stddev_pitch_dumbbell, var_pitch_dumbbell, avg_yaw_dumbbell, stddev_yaw_dumbbell, var_yaw_dumbbell, max_roll_forearm, max_picth_forearm, min_roll_forearm, min_pitch_forearm, amplitude_roll_forearm, amplitude_pitch_forearm, var_accel_forearm, avg_roll_forearm, stddev_roll_forearm, var_roll_forearm, avg_pitch_forearm, stddev_pitch_forearm, var_pitch_forearm, avg_yaw_forearm, stddev_yaw_forearm, var_yaw_forearm) )

```

Upon the remaining variables, it was helpful to perform some exploratory data analysis from here:
https://www.r-bloggers.com/how-to-use-data-analysis-for-machine-learning-example-part-1/

The melt chart of the remaining variables looks like this

```{r, warning=FALSE, message=FALSE}
Melt.PmlSub <- melt (PmlSub)
ggplot (data = Melt.PmlSub, aes (x=value)) + stat_density() + facet_wrap (~variable, scales = "free")
```

From this chart, these variables looked the least promising and were also removed:

amplitude_yaw_belt, var_yaw_belt, max_yaw_belt, min_yaw_belt, amplitude_roll_belt, amplitude_yaw_belt, stddev_yaw_belt, var_yaw_belt, max_yaw_dumbbell, min_yaw_dumbbell, amplitude_yaw_dumbbell, gyros_dumbbell_x, max_yaw_forearm, min_yaw_forearm, amplitude_yaw_forearm, gyros_forearm_y, gyros_forearm_z

After that, it was possible to train the random forest model on the remaining 48 variables. This is the code used to subset and train.

```{r, eval=FALSE}
library ("labeling", lib.loc="../../Documents/R/packages/")
library ("ggplot2", lib.loc="../../Documents/R/packages/")
library ("e1071", lib.loc="../../Documents/R/packages/")
library ("caret", lib.loc="../../Documents/R/packages/")

confusionMatrix (Test$classe, PredRf)
```

This is the confusion matrix showing the test set accuracy of the random forest model:

```{r}
confusionMatrix (Test$classe, PredRf)
```