---
output: html_document
---
# practicalmachinelearning
Repository for JHU Coursera Practical Machine Learning

The main issue in creating a model was that there were a very large number of variables, and thus building the more complex models like Random Forest took prohibitively long and sometimes produced warnings (e.g., "variable 'kurtosis_roll_belt' is not a factor").

Upon inspection, there were a number of variables that only assumed one of two possible values: blank and "div/0" (an undefined value resulting in a division by 0). These were:

kurtosis_yaw_belt, skewness_yaw_belt, kurtosis_yaw_dumbbell, skewness_yaw_dumbbell, amplitude_yaw_dumbbell, kurtosis_yaw_forearm, skewness_yaw_forearm, amplitude_yaw_forearm

```{r include=FALSE}
library ("labeling", lib.loc="../../Documents/R/packages/")
library ("ggplot2", lib.loc="../../Documents/R/packages/")
library ("e1071", lib.loc="../../Documents/R/packages/")
library ("caret", lib.loc="../../Documents/R/packages/")
load ("data/PmlSub48Rf.RData")
```

This is the confusion matrix showing the test set accuracy:

```{r}
confusionMatrix (Test$classe, PredRf)
```

Embed plots:

```{r, echo=FALSE}
plot (cars)
```
